---
title: "Assignment 2: Who Do We Know the Least About?"
subtitle: "Measuring uncertainty in American Community Survey data"
author: "Leo Leung"
date: today
format:
  html:
    toc: true
    code-fold: false
execute:
  warning: false
  message: false
---


## Setup

```{r}
library(tidyverse)
library(tidycensus)
```


---

## The question

Every number the American Community Survey publishes is an estimate drawn from a
sample, not a count. That means every number comes with a margin of error — a
statement about how wrong it might be.

Most people ignore it. In this assignment, you won't.


**In this analysis, I look at the median age across counties in Florida, and
ask which communities we measure most and least reliably.**

---

## Choosing what to measure



```{r}
my_state <- "FL"        # two-letter abbreviation, in quotes
my_variable <- "B01002_001"     # the variable code, in quotes
```

## Pulling the data

```{r}
county_data <- get_acs(
  geography = "county",
  variables = my_variable,
  state = my_state,
  year = 2023,
  survey = "acs5"
)
```

Let's look at what came back.

```{r}
dim(county_data)
head(county_data)
```


**[How many rows did you get? Does that match the number of counties in your state?]**

There are 67 counties in Florida and my row count matches.

---

## How big is the margin of error?

A margin of error of 3,000 means something very different for an estimate of 20,000
than for an estimate of 200,000. To compare across places, we need it as a
percentage.

```{r}
county_data <- county_data %>%
  mutate(moe_pct = moe / estimate * 100)
```


Now sort to see which counties are measured least reliably.

```{r}
county_data %>%
  arrange(desc(moe_pct)) %>%
  select(NAME, estimate, moe, moe_pct) %>%
  head(10)
```


**[Which county has the largest margin of error relative to its estimate? How large
is it?]**

Lafayette County, Florida has the largest margin of error relative to its estimate, which is around 6.22%.

---

## Sorting counties into reliability categories

```{r}
county_data <- county_data %>%
  mutate(reliability = case_when(
    moe_pct < 2  ~ "High confidence",
    moe_pct < 5 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))

count(county_data, reliability)
```


**[How many counties fall into each category?]**

If the cutoff is 5% and 10%, then 65 of them will land to high confidence and only 2 county will land to moderate. That is why I manually change the cutoffs to 2% and 5%, which showcase 59 in high confidence, 6 in moderate and 2 in low confidence.

---

## Is the margin of error bigger in small places?

To answer this we need population and our variable side by side, in the same row.
That means a **wide** pull.

```{r}
county_wide <- get_acs(
  geography = "county",
  variables = c(myvar = my_variable,
                pop   = "B01003_001"),
  state = my_state,
  year = 2023,
  survey = "acs5",
  output = "wide"
)

head(county_wide)
```


```{r}
county_wide <- county_wide %>%
  mutate(moe_pct = myvarM / myvarE * 100) %>%
  mutate(reliability = case_when(
    moe_pct < 2  ~ "High confidence",
    moe_pct < 5 ~ "Moderate",
    TRUE         ~ "Low confidence"
  ))
```

Now compare the average population of counties in each reliability category.

```{r}
county_wide %>%
  group_by(reliability) %>%
  summarize(
    n_counties = n(),
    avg_population = mean(popE)
  )
```


**[What is the average population of high-confidence counties versus low-confidence
counties? Is there a pattern?]**

The average population of high-confidence counties is 337126 in 65 counties, while the average population of low-confidence is 7842 in 2 counties. It means that larger population sizes have higher data confidence. 

---

## What this means for policy


**[Your 200–250 word reflection goes here.]**

1. In Florida, the communities we know the least about are small, rural counties such as Lafayette County and Liberty County. In our analysis, Lafayette County had the highest relative margin of error at 6.22%, while Liberty County, Florida have the margin of error at 5.37%. This two counties also fell into the moderate confidence category, with an average population of just 7,842.

2. This uncertainty stems directly from how the American Community Survey (ACS) operates. Because the ACS collects data from a sample of households rather than conducting a full census count, smaller populations naturally yield smaller sample sizes. With fewer completed responses, individual sample variations will lead to higher random variations on a bigger impact, driving up the margin of error percentage relative to the estimate.

3. If a state agency allocated funding based on median ages, like distributing senior services grants.  Treating every county's estimate as equally certain means that the rural residents would suffer most. In small counties, less sample calculated median age. If Lafayette County's median age were underestimated due to sampling error and there is more outliner, state algorithms might misclassify it as a younger community and withhold essential funding for senior centers, elder healthcare, and specialized transit. Meanwhile, large urban counties like Miami-Dade have massive sample sizes, negligible relative error, and stable funding streams. Relying uncritically on ACS estimates without accounting for margin of error disproportionately harms vulnerable, small-town communities by compounding statistical uncertainty with financial deprivation.

---

## Optional challenges - For those craving more


---

## Before you submit


